---
title: "Transient Oriented Development in Boston"
author: "Emily Zhou"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Version 1.0 \| First Created Sep 15, 2023 \| Updated Sep 20, 2023
Key words: Multiple Ring Buffer, Indicators, Transit Oriented Development, Gentrification, Boston

# Introduction

In city planning, transit-oriented development (TOD) is a type of urban development practice that aims to maximize the amount of residential, business, and leisure space within walking distance of public transport. TOD offers a range of benefits that all contribute to creating a more vibrant, sustainable, and inclusive communities. For example, a community with a strong and dependable transit system and streetscaping elements can discourage vehicle dependence and congestion. This further leads to reduction in greenhouse gas emissions per capita and a smaller carbon footprint. In addition, as cities become more walkable, TOD can stimulate economic growth by attracting business and creating jobs. This includes bringing in mixed-use developments, such as retail, dining, and entertainment options within easy reach of resident. Moreover, TOD increases property values in surrounding areas. For developers and governments, this land value capitalization is particularly important. If renters and home buyers are willing to pay more to live near transit amenities, it should be reflected in higher land values and property tax returns near stations. 

Transit oriented development has been a large part of Boston’s growth. The Boston Redevelopment Authority has been undertaking a TOD initiative to build on and reinforce Boston's existing and emerging neighborhood centers. Their goal is to **connect all neighborhood centers into a comprehensive transit network** and to **support that network with appropriate development patterns near these neighborhood centers**. To contribute to this work, this report conducts a comprehensive analysis of existing TOD patterns in Boston. Through visualizing temporal changes in demographic, social, and economic characteristics of neighborhoods across Boston, examining the spatial distribution of transit stops, and comparing the differences of development patterns between TOD and Non-TOD tracts, we aim to address the following questions: 

1. Who is living around transit stations in Boston? 

2. How does the social, economic, and demographic characteristics of households living near transit stations in Boston change over time? 

3. Are Bostonians willing to pay a premium to live in transit-rich areas? If residents value these locations, officials should consider changing the zoning code to allow increased density around transit.

4. How can planners in Boston be more strategic about developments in transit-rich neighborhoods in order to satisfy the needs of those who favor transit proximity? 

5. What are the limitations of this methodology to predict transit-oriented development? 

The GitHub repository for this report is available [here](https://github.com/emilyzhou112/MUSA5080-Boston-TOD). 

# Data Manipulation

Before we begin our analysis and visualization, there are a couple of data cleaning and manipulation steps that we need to perform as well as some analytical decisions to specify.  

```{r setup_packages, warning = FALSE, eval = FALSE, include=FALSE}

# list of required packages
packages <- c( "tidyverse", "ggplot2", "svDialogs", "tidycensus", "sf", "knitr", "rmarkdown", "kableExtra", "stringr", "biscale", "cowplot")

# install and load required packages
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE, quietly=TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# global options 
options(scipen=999)
options(tigris_class = "sf")
options(digits = 3)

# load in multiple ring buffer function
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

```


```{r load_key, warning = FALSE, eval = FALSE, message=FALSE, include=FALSE}

census_api_key(dlgInput(
  "Enter a Census API Key", # ask for an api key
  Sys.getenv("CENSUS_API_KEY")
)$res,
overwrite = TRUE)

```

## ACS Data

We are using the America Community Survey (5-year estimate) for year 2010 and 2019 in this report. We chose the ACS data because it collects detailed demographic, social, economic, and housing data on a wide range of topics, providing a more comprehensive view of a community's characteristics in Boston. We chose 2019 as the end year because COVID-19 since 2020 has drastically changed people's work and travel behaviors: the widespread adoption of remote work during the pandemic would certainly led to decline in the demand for office space and the need for transit-oriented office developments. In order to better understand if there's a difference between individuals living in TOD and Non-TOD neighborhood over time, we chose the following ACS variables, categorized into demographic, social, and economic characteristics, for our analysis. 

| Variable Name              | ACS Name    | 
| :------------------------: | :----------:|  
| **Demographic**            |             |
| Total Population           | B02001_001  |
| White Population           | B02001_002  |
| Black Population           | B02001_003  |
| Asian Population           | B02001_005  |
| Latinx Population          | B03002_012  |
| Male 70-74                 | B01001_022  |
| Male 75-79                 | B01001_023  |
| Male 80-84                 | B01001_024  |
| Male over 85               | B01001_025  |
| Female 70-74               | B01001_046  |
| Female 75-79               | B01001_047  |
| Female 80-84               | B01001_048  |
| Female over 85             | B01001_049  |
| Non-US Citizen             | B05001_006  |
| **Economic**               |             |
| Median Household Income    | B19013_001  |
| Median Rent                | B25058_001  |
| Population under Poverty   | B06012_002  |
| **Social**                 |             |
| Bachelor Degree            | B06009_005  |
| Graduate Degree            | B06009_006  |
| Renter Occupied Household  | B25026_009  |
| Population without Vehicle | B08014_002  |

Based on these ACS variables, we further computed the following variables:
- **PctMinority** : percentage of minority population, where minority here refers to anyone who is Asian, Latinx, or Black.   
- **PctAging**: percentage of aging population, where aging here refers to anyone who is over 70 years old.  
- **PctForeign**: percentage of population who are non-US citizens.  
- **PctBachelors**: percentage of population with at least a Bachelor's degree.   
- **PctRenter**: percentage of renter occupied households.   
- **PctNoCar**:percentage of population without a vehicle.  
- **PctPoverty**: poverty rate.  

The following code gives an example of the data cleaning steps we performed using 2010 ACS data, where we downloaded the ACS data, reprojected the data, removed census tracts outside of Boston, renamed the columns, performed additional data manipulations, and dumped unnecessary variables. 

```{r define_acs_var, warning = FALSE, eval = FALSE, message=FALSE, include=FALSE}

acs_var = c("B02001_001E", # total population
            "B02001_002E", # white population
            "B02001_003E", # black population
            "B02001_005E", # asian population
            "B03002_012E", # latinx population
            "B01001_022E", # male 70-74
            "B01001_023E", # male 75-79
            "B01001_024E", # male 80-85
            "B01001_025E", # male over 85
            "B01001_046E", # female 70-74
            "B01001_047E", # female 75-79
            "B01001_048E", # female 80-85
            "B01001_049E", # female over 85
            "B19013_001E", # median household income
            "B25058_001E", # median rent
            "B06012_002E", # total poverty
            "B06009_005E", # bachelor
            "B06009_006E", # graduate
            "B08014_002E", # no car
            "B25026_009E", # renter
            "B05001_006E" # non-us citizen
            )

```


```{r non_boston_county, warning = FALSE, eval = FALSE, message=FALSE, include=FALSE}

remove <- c("25025180400", "25025990101", "25025180301", "25025160400", "25025180500", "25025160601", "25025180200", "25025160300", "25025160200", "25025180101", "25025170502", "25025160502", "25025170701", "25025170501", "25025170800", "25025170200", "25025170400", "25025170100", "25025170100", "25025170601", "25025170300", "25025170702","25025160602", "25025160501", "25025160101")

```


```{r boston_2010_data, warning = FALSE, eval = FALSE, message=FALSE,  class.source = 'fold-show'}

boston10 <- 
  get_acs(geography = "tract", variables = acs_var, 
          year=2010, state='MA', county=025, geometry=T, output="wide") %>% 
  st_transform('EPSG:26986') %>% 
  filter(!(GEOID %in% remove)) %>% 
  rename(TotalPop = B02001_001E, 
         Whites = B02001_002E,
         African_Americans = B02001_003E,
         Asians = B02001_005E,
         Latinx = B03002_012E,
         M70_74 = B01001_022E,
         M75_79 = B01001_023E,
         M80_85 = B01001_024E,
         M85 = B01001_025E,
         FM70_74 = B01001_046E,
         FM75_79 = B01001_047E,
         FM80_85 = B01001_048E,
         FM85 = B01001_049E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         College = B06009_005E,
         Graduate = B06009_006E,
         NoCar = B08014_002E,
         Renter = B25026_009E,
         Foreigner = B05001_006E
         ) %>% 
  dplyr::select(-starts_with("B"))%>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctMinority = ifelse(TotalPop > 0, (African_Americans + Asians + Latinx ) / TotalPop, 0),
         pctAging = ifelse(TotalPop > 0, (M70_74 + M75_79 + M80_85 + M85 + FM70_74 + FM75_79 + FM80_85 + FM85 ) / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((College + Graduate) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctNoCar = ifelse(TotalPop > 0, NoCar / TotalPop, 0),
         pctRenter = ifelse(TotalPop > 0, Renter / TotalPop, 0),
         pctForeigner = ifelse(TotalPop > 0, Foreigner / TotalPop, 0),
         year = "2010") %>% 
  dplyr::select(-Whites,-African_Americans, -Asians, -Latinx, -M70_74, -M75_79, -M80_85, -M85, -FM70_74, -FM75_79, -FM80_85, -FM85, -College, -Graduate, -TotalPoverty, -NoCar, -Renter, -Foreigner)

```

Note: Boston city is part of the larger Suffolk county in Massachusetts, which also contains the city of Chelsea, Revere, and Winthrop. To limit our analysis within Boston, we have to remove census tracts outside. The map below shows Boston relative to other cities in Suffolk county.  

```{r suffolk_county, warning = FALSE, eval = FALSE, message=FALSE}

Suffolk <- 
  get_acs(geography = "tract", variables = "B02001_001E", 
          year=2010, state='MA', county=025, geometry=T, output="wide")

ggplot() + 
  geom_sf(data=st_union(Suffolk), fill= "#D3D3D3", color = "white", size = 2) +
  geom_sf(data=st_union(boston10), fill= "#bd7ebe", color = "white", size = 2) +
  labs(title = "Boston City in Suffolk County",
       caption = "Data: American Community Survey 2010 and 2019",
       color = "Boston") +
  theme_light() +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  annotate(geom = "text",x = -71.10,y = 42.3, size = 4, label = "Boston",color = "white") +
  annotate(geom = "text",x = -71.04,y = 42.40, size = 3, label = "Chelsea",color = "black") +
  annotate(geom = "text",x = -71.01,y = 42.42, size = 3, label = "Revere",color = "black") +
  annotate(geom = "text",x = -70.97,y = 42.38, size = 3, label = "Winthrop",color = "black") 

```

Now, we repeat the same process to obtain the 2019 ACS data and combine it with the 2010 data. 

```{r boston_2019_data, warning = FALSE, eval = FALSE, message=FALSE}

boston19 <- 
  get_acs(geography = "tract", variables = acs_var, 
          year=2019, state='MA', county=025, geometry=T, output="wide") %>% 
  st_transform('EPSG:26986') %>% 
  filter(!(GEOID %in% remove)) %>% 
  rename(TotalPop = B02001_001E, 
         Whites = B02001_002E,
         African_Americans = B02001_003E,
         Asians = B02001_005E,
         Latinx = B03002_012E,
         M70_74 = B01001_022E,
         M75_79 = B01001_023E,
         M80_85 = B01001_024E,
         M85 = B01001_025E,
         FM70_74 = B01001_046E,
         FM75_79 = B01001_047E,
         FM80_85 = B01001_048E,
         FM85 = B01001_049E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         College = B06009_005E,
         Graduate = B06009_006E,
         NoCar = B08014_002E,
         Renter = B25026_009E,
         Foreigner = B05001_006E
         ) %>% 
  dplyr::select(-starts_with("B")) %>% 
    mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctMinority = ifelse(TotalPop > 0, (African_Americans + Asians + Latinx ) / TotalPop, 0),
         pctAging = ifelse(TotalPop > 0, (M70_74 + M75_79 + M80_85 + M85 + FM70_74 + FM75_79 + FM80_85 + FM85 ) / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((College + Graduate) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctNoCar = ifelse(TotalPop > 0, NoCar / TotalPop, 0),
         pctRenter = ifelse(TotalPop > 0, Renter / TotalPop, 0),
         pctForeigner = ifelse(TotalPop > 0, Foreigner / TotalPop, 0),
         year = "2019") %>% 
  dplyr::select(-Whites,-African_Americans, -Asians, -Latinx, -M70_74, -M75_79, -M80_85, -M85, -FM70_74, -FM75_79, -FM80_85, -FM85, -College, -Graduate, -TotalPoverty, -NoCar, -Renter, -Foreigner)

allTracts <- rbind(boston10,boston19)
```


## Transit Data

The Boston transit data is provided by [MassGIS](https://www.mass.gov/info-details/massgis-data-mbta-rapid-transit#overview-), through which we downloaded the raw file and uploaded it to our GitHub repository. To ensure data consistency, we have repojected this layer for it to match with our census data, and have re-categorized transit lines. 


```{r transit_data, warning = FALSE, eval = FALSE, message=FALSE, class.source = 'fold-show'}

boston_stops <- st_read("https://raw.githubusercontent.com/emilyzhou112/MUSA5080-Boston-TOD/main/data/boston_stps.geojson") %>%
  st_transform(st_crs(boston10)) %>% 
  mutate(stops = ifelse(grepl("/", LINE), "OTHER", LINE)) 


```

The MBTA Rapid Transit layers represent the core subway system lines and stations within the Greater Boston area. There are five subways lines in Boston: Red, Green, Orange, Blue, and Silver, which services many Boston neighborhoods plus parts of Cambridge, Somerville, Revere, Malden, Brookline, Newton, Milton, and Quincy. To limit our scope of analysis within Boston, we have removed subway stations outside of Boston. The map below shows transit stops used in our study as well as those that we have excluded. 

```{r visualize_stops, warning = FALSE, eval = FALSE, message=FALSE}

MBTA <- st_read("https://raw.githubusercontent.com/emilyzhou112/MUSA5080-Boston-TOD/main/data/MBTA_node.geojson") %>%
  st_transform(st_crs(boston10))

ggplot() + 
  geom_sf(data=st_union(boston10), fill= "#D3D3D3", color = "white", size = 2) +
  geom_sf(data=MBTA, color = "black", size= 1) + 
  geom_sf(data=boston_stops, 
          aes(colour = stops %>% str_to_title()),
          show.legend = "point", size= 1.5) + 
  scale_colour_manual(values = c("#7eb0d5","#b2e061", "#ffb55a", "#ffee65", "#fd7f6f", "#bd7ebe")) +
  labs(title = "Boston MBTA Rapid Transit Stops",
       caption = "Data: MassGIS 2022",
       colour = "Lines") +
  theme_light() +
  theme(plot.title = element_text(size = 12, face = "bold"))

```

To delineate census tracts that are in close proximity to transit stations, we created a half mile buffer around all transit stations, and compared it to all the centroids of our census tracts. If the centroid is located within the half mile buffer, we categorize that tract as "TOD"; otherwise, this tract is considered "Non-TOD". An additional data manipulation here is to adjust median rent and median household income for inflation.  

```{r delineate_TOD, warning = FALSE, eval = FALSE, message=FALSE, class.source = 'fold-show'}

stops_buffer <- st_union(st_buffer(boston_stops, 804)) %>% 
      st_sf() 

allTracts.group <- 
  rbind(
    st_centroid(allTracts)[stops_buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[stops_buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2010", MedRent * 1.17, MedRent)) %>% 
  mutate(MedHHInc.inf = ifelse(year == "2010", MedHHInc * 1.17, MedHHInc)) 

```

The map below shows the distribution of TOD and Non-TOD tracts in Boston, using the above method. 

```{r visualize_tod,  warning = FALSE, eval = FALSE, message=FALSE}

allTracts.group %>% 
  filter(year == 2010 ) %>% 
  ggplot() + 
  geom_sf(aes(fill = TOD), color = "transparent") +
  scale_fill_manual(values = c("#7eb0d5", "#b2e061")) + 
  geom_sf(data=boston_stops, 
          color = "#fd7f6f", size= 0.8)+
  labs(title = "Transient Oriented Development Tracts",
       subtitle = "Tracts within 0.5 mile of transit stops",
       caption = "Data: American Community Survey 2010 and 2019",
       fill = "Tracts Types") +
  theme_light() +   
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"))


```

# Data Analysis

## Demographic Characteristics 


```{r percentage_aging, warning = FALSE, eval = FALSE, message=FALSE}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctAging), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "blue4", name= "Percentage Aging", breaks = c(0, 0.1, 0.2, 0.3))+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
  theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage of Aging Population in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 
 

```


```{r percent_foreigner}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctForeigner), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "blue4", name= "Percentage Foreigner", breaks = c(0, 0.2, 0.4, 0.6))+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
 theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage of Non-US Citizens in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 
 
```

```{r percent_minorities}
allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctMinority), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "blue4", name= "Percentage Minority", breaks = c(0, 0.3, 0.6, 0.9))+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
 theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage of Minorities in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 
```


## Economic Characteristics

```{r median_rent}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=MedRent.inf), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "seagreen4", name= "Rent")+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
   theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Median Rent in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops \n2019 data adjusted for inflation",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 

```
```{r poverty}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctPoverty), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "seagreen4", name= "Percentage Poverty")+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
 theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage Poverty in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022")

```


```{r}

allTracts.group %>% 
  mutate(inc = MedHHInc.inf/1000) %>% 
  ggplot()+
  geom_sf(aes(fill=inc), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "seagreen4", name= "Income (thousands)")+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
  theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Median Household Income in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops\n2019 data adjusted for inflation",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022")

```



## Social Characteristics 

```{r renters}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctRenter), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "#0047AB", name= "Percentage Renters")+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
   theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage Renters in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 

```


```{r no car}

allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctNoCar), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "#0047AB", name= "Percentage No Vehicle", breaks = c(0, 0.2, 0.4, 0.6))+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
   theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage of Population without Car in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 


```

```{r}
allTracts.group %>% 
  ggplot()+
  geom_sf(aes(fill=pctBachelors), color="transparent") +
  scale_fill_continuous(low = "#FAF9F6", high = "#0047AB", name= "Percentage Bachelors", breaks = c(0, 0.3, 0.6, 0.9))+
  facet_wrap(~year) +
  geom_sf(data=stops_buffer, color = "red", fill = "transparent", linewidth = 0.4) +
   theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) + 
  labs(title = "Percentage of Population with at Least a Bacehlor's Degree in Boston",
       subtitle = "Red circles indicate areas within 0.5 mile of a transit stops",
       caption = "Data: American Community Survey 2010 and 2019 \n MassGIS 2022") 

```

## Overall Analysis 

```{r all_variables_summarize}

allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent.inf, na.rm = T),
            Income = mean(MedHHInc.inf, na.rm = T),
            Pop = mean(TotalPop, na.rm = T),
            PctMinority = mean(pctMinority, na.rm = T),
            PctAging = mean(pctAging, na.rm = T),
            PctBach = mean(pctBachelors, na.rm = T),
            PctPoverty = mean(pctPoverty, na.rm = T),
            PctNoCar = mean(pctNoCar, na.rm = T),
            PctRenter = mean(pctRenter, na.rm = T),
            PctForeign = mean(pctForeigner, na.rm = T))

```


```{r all_variables_table}

kable(allTracts.Summary, format = "html") %>%
 kable_styling(full_width = T, 
               bootstrap_options = c("striped", "hover"),
               font_size = 12) %>% 
  footnote(general_title = "\n",
           general = "Table 1: Comparaing Population Characteristics Across Space and Time ") %>% 
  row_spec(0, font_size = 14) %>% 
  row_spec(1, color = "white", background = "#7eb0d5") %>% 
  row_spec(2, color = "white", background = "#b2e061") %>% 
  row_spec(3, color = "white", background = "#7eb0d5") %>% 
  row_spec(4, color = "white", background = "#b2e061") %>% 
  column_spec(1:2, bold = T)

# 

```


```{r all_variables_plot}

allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=5) +
  scale_fill_manual(values = c("#7eb0d5", "#b2e061")) +  
  labs(title = "Indicator Differences Across Time and Space",
       caption = "Data: American Community Survey 2010 and 2019") +
  theme(legend.position = "bottom",
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) 
```



```{r percentage_change}

allTracts.Long <- allTracts.Summary %>%
  pivot_longer(cols = c(Rent, Income, Pop, PctMinority, PctAging, PctBach, PctPoverty, PctNoCar, PctRenter, PctForeign), names_to = "Variable", values_to = "Value") %>%
  pivot_wider(names_from = year, values_from = Value, names_prefix = "Year") %>% 
  mutate(pct_change = (Year2019-Year2010) / Year2010)

```


```{r all_variables_bar}

allTracts.Long %>% 
  ggplot() +
  geom_bar(aes(x = Variable, y = pct_change, fill = TOD),
           stat = "identity") +
  facet_wrap(~TOD) +
  coord_flip() +
  scale_fill_manual(values = c("#7eb0d5", "#b2e061")) +  
  labs(title = "Percentage Change 2010-2019 for TOD and Non-TOD Tracks",
       caption = "Data: American Community Survey 2010 and 2019") +
  ylab ("Percentage Change") +
  theme(legend.position = "bottom",
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) 
  


```

Observations: 
- Rent in Boston has been increasing, but the increase in TOD is more salient. 
- Population in Boston has been increasing, but the increase in non-TOD is more salient
- Renters stay pretty much the same throughout. 
- Poverty rate greatly reduced, especially in TOD areas. 
- More people in Non-TOD tracks do not have car. 
- More minority population starts to live in TOD areas. 
- Education attainment in Boston greatly increase in both TOD and non-TOD. 
- Percentage Aging population starts to live in TOD.
- Income increases in both TOD and non-TOD. 

## Further Analysis 

```{r graduated_prep}

boston_stops$ID <- seq_along(boston_stops$STATION)

stop_attributes <- st_buffer(boston_stops, 804) %>% 
  st_sf() %>% 
  st_intersection(st_centroid(boston10)) %>% 
  st_drop_geometry() %>% 
  group_by(ID) %>% 
  summarize(
    total_population = sum(TotalPop),
    median_rent = mean(MedRent, na.rm = T))

stops_summary <- boston_stops %>% 
  left_join(stop_attributes, by = "ID")

```

```{r graduated_pop, warning=FALSE}

ggplot() +
  geom_sf(data=st_union(boston10), fill= "#D3D3D3", color = "white", size = 2) +
  geom_sf(data=stops_summary, aes(size=total_population, color = stops %>% str_to_title())) +
  scale_size(range = c(0.5, 3), name = "Population") +    
  scale_colour_manual(values = c("#7eb0d5","#b2e061", "#ffb55a", "#ffee65", "#fd7f6f", "#bd7ebe"), name = "Stops") +
  labs(title = "Population within 0.5 Mile of Tansit Stops",
       subtitle = "Dot size proportional to population",
       caption = "Data: American Community Survey 2010 and 2019") +
  theme_light() +   
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"))
```

```{r graduated_rent, warning=FALSE}


ggplot() +
  geom_sf(data=st_union(boston10), fill= "#D3D3D3", color = "white", size = 2) +
  geom_sf(data=stops_summary, aes(size=median_rent, color = stops %>% str_to_title())) +
  scale_size(range = c(0.5, 3), name = "Rent") +    
  scale_colour_manual(values = c("#7eb0d5","#b2e061", "#ffb55a", "#ffee65", "#fd7f6f", "#bd7ebe"), name = "Stops") +
  labs(title = "Median Rent within 0.5 Mile of Tansit Stops",
       subtitle = "Dot size proportional to rent",
       caption = "Data: American Community Survey 2010 and 2019") +
  theme_light() +   
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"))
```

```{r boston_MRB}

boston_MRB <- multipleRingBuffer(st_union(boston_stops), 14484, 804)
ggplot() +
    geom_sf(data=boston_MRB, aes(fill = distance), color = "#708090") +
    scale_fill_continuous(low = "#FAF9F6", high = "#beb9db", name= "Distance")+
    geom_sf(data=st_union(boston10), fill= "#D3D3D3", color = "white", size = 2) +
    geom_sf(data=boston_stops, 
          color = "#fd7f6f", size= 0.8) +    
    labs(title="Half Mile Buffers",
         subtitle = "Distance in-between each buffer is 0.5 mile",
         caption = "Data: American Community Survey 2010 and 2019 \nMassGIS 2022") +
    theme_light() +   
    theme(plot.subtitle = element_text(size = 9,face = "italic"),
          plot.title = element_text(size = 12, face = "bold"))

```

```{r line_plot}

allTracts.rings <-
  st_join(st_centroid(dplyr::select(allTracts, GEOID, year)),
          boston_MRB) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(allTracts, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() 


allTracts.rings.summary <- st_drop_geometry(allTracts.rings) %>%
  group_by(distance, year) %>%
  summarize(Mean_Rent = mean(MedRent, na.rm=T))


ggplot(allTracts.rings.summary, aes(x = distance, y = Mean_Rent)) + 
  geom_line(aes(color = year), size = 2) + 
  scale_color_manual(values = c("#7eb0d5", "#b2e061"), name = "Year") +
  labs(title = "Mean Rent as a Function of Distance to Subway Stations",
       caption = "Data: American Community Survey 2010 and 2019") +
  xlab ("Distance from Transit Stops") +
  ylab("Mean Rent") + 
  theme(legend.position = "bottom",
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        ) 

```

## Bivariate informing planning decisions 


```{r bivariate_map}


bi_data <- bi_class(boston19, x = pctMinority, y = pctAging, style = "quantile", dim = 3) #  a 3*3 color-schemed bi-variate choropleth 

bi_map <- ggplot() +
  geom_sf(data = bi_data, mapping = aes(fill = bi_class), color = "transparent", show.legend = FALSE) +
  bi_scale_fill(pal = "DkCyan", dim = 3) + 
  labs(title = "Pct Minority, Pct Aging Population") + 
  geom_sf(data = boston_stops, colour = "red", alpha = 0.8, size = 0.5) +
  theme_light() +
  theme(plot.title = element_text(size = 12, face = "bold"))

legend <- bi_legend(pal = "DkCyan",
                    dim = 3,
                    xlab = "PctMinority ",
                    ylab = "PctAging ",
                    size = 5.8)
  
ggdraw() +
  draw_plot(bi_map, 0, 0, 1, 1) +
  draw_plot(legend, 0.1, 0.1, 1.2, 0.23)
```







